---
title: "Data visualization lab"
author: "Marco Chierici"
date: "Mar 8, 2023"
output:
  pdf_document:
    toc: no
  html_document:
    theme: readable
    toc: no
    toc_float: yes
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 60),
  tidy = TRUE,
  error = TRUE,
  collapse = TRUE,
  comment = "#>"
)
set.seed(4561)
```

# Data handling in R

(partially abridged from Jenny Bryan, see License information at the end)

## Before we start

### RStudio essential shortcuts

Remember some of the most commonly used RStudio shortcuts:

-   function or dataset help:
    -   press <kbd>F1</kbd> with your cursor anywhere in a function name; or
    -   type `?function_name` in the console
-   execute from script:
    -   <kbd>Ctrl</kbd> + <kbd>Enter</kbd> (Windows/Linux)
    -   <kbd>Cmd</kbd> + <kbd>Enter</kbd> (macOS)
-   assignment operator (`<-`):
    -   <kbd>Alt</kbd> + <kbd>-</kbd> (Windows/Linux)
    -   <kbd>Option</kbd> + <kbd>-</kbd> (macOS)
-   pipe operator (`%>%`):
    -   <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>m</kbd> (Windows/Linux)
    -   <kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>m</kbd> (macOS)
-   insert code chunk in Rmd:
    -   <kbd>Ctrl</kbd> + <kbd>Alt</kbd> + <kbd>i</kbd> (Windows/Linux)
    -   <kbd>Cmd</kbd> + <kbd>Option</kbd> + <kbd>i</kbd> (macOS)

## Data frames are awesome

Whenever you have rectangular, spreadsheet-y data, your default data receptacle in R is a data frame. Do not depart from this without good reason. Data frames are awesome because...

-   Data frames package related variables neatly together,
    -   keeping them in sync vis-a-vis row order
    -   applying any filtering of observations uniformly
-   Most functions for inference, modelling, and graphing are happy to be passed a data frame via a `data =` argument. This has been true in base R for a long time.
-   The set of packages known as the [tidyverse](https://tidyverse.tidyverse.org/) takes this one step further and explicitly prioritizes the processing of data frames. This includes popular packages like dplyr and ggplot2. In fact the tidyverse prioritizes a special flavor of data frame, called a "tibble".

Data frames -- unlike general arrays or, specifically, matrices in R -- can hold variables of different flavors, such as character data (subject ID or name), quantitative data (white blood cell count), and categorical information (treated vs. untreated). If you use homogeneous structures, like matrices, for data analysis, you are likely to make the terrible mistake of spreading a dataset out over multiple, unlinked objects. Why? Because you can't put character data, such as subject name, into the numeric matrix that holds white blood cell count. This fragmentation is a Bad Idea.

## Get the Gapminder data

We will work with some of the data from the [Gapminder project](https://www.gapminder.org/), which was also released as an [R package](https://cran.r-project.org/package=gapminder). We can install it from CRAN like so:

```{r eval = FALSE}
install.packages("gapminder")
```

Now load the package:

```{r}
library(gapminder)
```

### Installing packages

Here's a quick tip about installing packages. Say you want to install a packages on the fly only if it is not already in your library:

```{r, eval=FALSE}
if (!require(gapminder)) {
  install.packages("gapminder")
  library(gapminder)
}
```

`require()` is the same as `library()`, but it also outputs a `TRUE/FALSE` if the package is installed or not.

This can be further improved with the `pacman` (**pac**kage **man**ager) library! It simplifies the process of installing and loading libraries. First, we install and load it the usual way:

```{r, eval=FALSE}
install.packages("pacman")
library(pacman)
```

Now, loading or installing a package on the fly is as easy as:

```{r, eval=FALSE}
p_load(gapminder)
```

The function is smart enough to download, install, and load the package if it is not in your library already. Note how the package name does not need to be a string. The beauty of `pacman` is that you can also load multiple libraries at once:

```{r, eval=FALSE}
# the old way
library(tidyverse)
library(gapminder)
library(openxlsx)
#library(yetAnotherLibrary)

# the new way
p_load(tidyverse, gapminder, openxlsx, yetAnotherLibrary)
```

## Meet the `gapminder` data frame, or "tibble"

By loading the gapminder package, we now have access to a data frame by the same name. Get an overview of this with `str()`, which displays the structure of an object.

```{r message = FALSE, warning = FALSE, live=TRUE}
str(gapminder)
```

`str()` will provide a sensible description of almost anything and, worst case, nothing bad can actually happen. When in doubt, just `str()` some of the recently created objects to get some ideas about what to do next.

We could print the `gapminder` object itself to screen. However, if you've used R before, you might be reluctant to do this, because large datasets just fill up your Console and provide very little insight.

This is the first big win for **tibbles**. The [tidyverse](https://tidyverse.tidyverse.org/) offers a special case of R's default data frame: the "tibble", which is a nod to the actual class of these objects, `tbl_df`.

If you have not already done so, install the tidyverse meta-package now:

```{r eval = FALSE}
install.packages("tidyverse")
```

Now load it:

```{r}
library(tidyverse)
```

Now we can boldly print `gapminder` to screen! It is a tibble (and also a regular data frame) and the tidyverse provides a nice print method that shows the most important stuff and doesn't fill up your Console.

```{r, live=TRUE}
## see? it's still a regular data frame, but also a tibble
class(gapminder)
gapminder
```

If you are dealing with plain vanilla data frames, you can rein in data frame printing explicitly with `head()` and `tail()`. Or turn it into a tibble with `as_tibble()`!

```{r, live=TRUE}
head(gapminder)
tail(gapminder)
as_tibble(iris)
```

More ways to query basic info on a data frame:

```{r, live=TRUE}
names(gapminder)
ncol(gapminder)
length(gapminder)
dim(gapminder)
nrow(gapminder)
```

A statistical overview can be obtained with `summary()`:

```{r}
summary(gapminder)
```

Although we haven't begun our formal coverage of visualization yet, it's so important for smell-testing a dataset that we will make a few figures anyway. Here we use only base R graphics, which are... well, very basic.

```{r first-plots-base-R, live=TRUE}
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)
plot(lifeExp ~ log(gdpPercap), gapminder)
```

Let's go back to the result of `str()` to talk about what a data frame is.

```{r}
str(gapminder)
```

A data frame is a special case of a *list*, which is used in R to hold just about anything. Data frames are a special case where the length of each list component is the same. Data frames are superior to matrices in R because they can hold vectors of different flavors, e.g. numeric, character, and categorical data can be stored together. This comes up a lot!

## Look at the variables inside a data frame

To specify a single variable from a data frame, use the dollar sign `$`. Let's explore the numeric variable for life expectancy.

```{r histogram-lifeExp}
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

The year variable is an integer variable, but since there are so few unique values it also functions a bit like a categorical variable.

```{r}
summary(gapminder$year)
table(gapminder$year)
```

The variables for country and continent hold truly categorical information, which is stored as a *factor* in R.

```{r}
class(gapminder$continent)
summary(gapminder$continent)
```

```{r, live=TRUE}
levels(gapminder$continent)
nlevels(gapminder$continent)
```

The **levels** of the factor `continent` are "Africa", "Americas", etc. and this is what's usually presented to your eyeballs by R. In general, the levels are friendly human-readable character strings, like "male/female" and "control/treated". But *never ever ever* forget that, under the hood, R is really storing integer codes 1, 2, 3, etc. Look at the result from `str(gapminder$continent)` if you are skeptical.

```{r}
str(gapminder$continent)
```

This [Janus](https://en.wikipedia.org/wiki/Janus)-like nature of factors means they are rich with booby traps for the unsuspecting but they are a necessary evil. Learning how to properly care and feed for factors is crucial in modelling and figure-making.

Here we count how many observations are associated with each continent and, as usual, try to portray that info visually. This makes it much easier to quickly see that African countries are well represented in this dataset.

```{r tabulate-continent, live=TRUE}
table(gapminder$continent)
barplot(table(gapminder$continent))
```

## Exercises

Let's get the data for just 2007. How many rows? How many observations per continent?

```{r, live=TRUE}
hDat <- subset(gapminder, subset = year == 2007)
str(hDat)
table(hDat$continent)
```

If you want just some rows and/or just some variables, for inspection or to assign as a new object, use `subset()`:

```{r subset, live=TRUE}
# subset on country "Cambodia"
subset(gapminder, subset = country == "Cambodia")
# subset on country "Japan" or "Belgium"
subset(gapminder, subset = country %in% c("Japan", "Belgium"))
# subset on year 1952
subset(gapminder, subset = year == 1952)
# subset on "Uruguay", also selecting only the country, year, and lifeExp columns
subset(gapminder, subset = country == "Uruguay", select = c(country, year, lifeExp))

# plot lifeExp vs year only for "Zimbabwe"
plot(lifeExp ~ year, gapminder, subset = country == "Zimbabwe")
# plot lifeExp vs log(gdpPercap) only for 2007
plot(lifeExp ~ log(gdpPercap), gapminder, subset = year == 2007)
```

Now get data for which life expectancy is less than 32 years, assign it to an object, and determine how many observations are there overall and by continent.

```{r ex, live=TRUE}
df_sub <- subset(gapminder, subset = lifeExp < 32)
nrow(df_sub)
table(df_sub$continent)
```

------------------------------------------------------------------------

# Transforming data with `dplyr`

## Intro

[dplyr](https://dplyr.tidyverse.org/) is a package for data manipulation, developed by Hadley Wickham and Romain Francois. It is built to be fast, highly expressive, and open-minded about how your data is stored. It is installed as part of the [tidyverse](https://tidyverse.tidyverse.org/) meta-package and, as a core package, it is among those loaded via `library(tidyverse)`.

dplyr's roots are in an earlier package called [plyr](http://plyr.had.co.nz/), which implements the ["split-apply-combine" strategy for data analysis](https://www.jstatsoft.org/article/view/v040i01). Where plyr covers a diverse set of inputs and outputs (e.g., arrays, data frames, lists), dplyr has a laser-like focus on data frames or, in the tidyverse, "tibbles". dplyr is a package-level treatment of the `ddply()` function from plyr, because "data frame in, data frame out" proved to be so incredibly important.

Have no idea what I'm talking about? Not sure if you care? If you use these base R functions: `subset()`, `apply()`, `[sl]apply()`, `tapply()`, `aggregate()`, `split()`, `do.call()`, `with()`, `within()`, then you should keep reading. Also, if you use `for()` loops a lot, you might enjoy learning other ways to iterate over rows or groups of rows or variables in a data frame.

## Think before you create excerpts of your data ...

If you feel the urge to store a little snippet of your data:

```{r}
(canada <- gapminder[241:252, ])
```

Stop and ask yourself ...

> Do I want to create mini datasets for each level of some factor (or unique combination of several factors) ... in order to compute or graph something?

If YES, **use proper data aggregation techniques** or "faceting" in `ggplot2` (we'll cover this in a later lesson!) -- **don't subset the data**. Or, more realistic, only subset the data as a temporary measure while you develop your elegant code for computing on or visualizing these data subsets.

Seriously consider whether you can achieve your goals by simply using the `subset =` argument of, e.g., the `lm()` function, to limit computation to your excerpt of choice. Lots of functions offer a `subset =` argument!

Copies and excerpts of your data clutter your workspace, invite mistakes, and sow general confusion. Avoid whenever possible.

Reality can also lie somewhere in between. You will find the workflows presented below can help you accomplish your goals with minimal creation of temporary, intermediate objects.

## Use `filter()` to subset data row-wise

`filter()` takes logical expressions and returns the rows for which all are `TRUE`.

```{r, live=TRUE}
# select rows where lifeExp < 29
filter(gapminder, country == "Russia")
# select data for Rwanda after 1979
filter(gapminder, country == "Rwanda", year > 1979)
# select data for Rwanda or Afghanistan (version 1)
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
# select data for Rwanda or Afghanistan (version 2)
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

Compare with some base R code to accomplish the same things:

```{r eval = FALSE}
gapminder[gapminder$lifeExp < 29, ] ## repeat `gapminder`, [i, j] indexing is distracting
subset(gapminder, country == "Rwanda") ## almost same as filter; quite nice actually
```

Under no circumstances should you subset your data the way I did before:

```{r eval = FALSE}
excerpt <- gapminder[241:252, ]
```

Why is this a terrible idea?

-   It is **not self-documenting**. What is so special about rows 241 through 252?
-   It is **fragile**. This line of code will produce different results if someone changes the row order of `gapminder`, e.g. sorts the data earlier in the script.

```{r eval = FALSE}
filter(gapminder, country == "Canada")
```

This call explains itself and is fairly robust.

## Meet the pipe operator

Before we go any further, we should exploit the new pipe operator that the tidyverse imports from the [magrittr](https://magrittr.tidyverse.org/) package by Stefan Bache. This is going to change your data analytical life. You no longer need to enact multi-operation commands by nesting them inside each other. This new syntax leads to code that is much easier to write and to read.

Here's what it looks like: `%>%` (see above for convenient keyboard shortcuts).

```{r, live=TRUE}
gapminder %>% head()
```

This is equivalent to `head(gapminder)`. The pipe operator takes the thing on the left-hand-side and **pipes** it into the function call on the right-hand-side -- literally, drops it in as the first argument.

Never fear, you can still specify other arguments to this function! To see the first 3 rows of `gapminder`, we could say `head(gapminder, 3)` or this:

```{r, live=TRUE}
gapminder %>% head(3)
```

If the assignment operator, `<-`, should make you think "gets", similarly you should think "then" whenever you see the pipe operator, `%>%`.

You are probably not impressed yet, but the magic will soon happen.

## Use `select()` to subset the data on variables or columns.

Back to dplyr...

Use `select()` to subset the data on variables or columns. Here's a conventional call for selecting the columns `year` and `lifeExp`:

```{r, live=TRUE}
select(gapminder, year, lifeExp)
```

And here's the same operation, but written with the pipe operator and piped through `head()`:

```{r, live=TRUE}
gapminder %>%
  select(year, lifeExp) %>%
  head(4)
```

Think: "Take `gapminder`, then select the variables year and lifeExp, then show the first 4 rows."

## Revel in the convenience

Here's the data for Cambodia, but only for certain variables (`year`, `lifeExp`):

```{r, live=TRUE}
gapminder %>%
  filter(country == "Cambodia") %>%
  select(year, lifeExp)
```

and what a typical base R call would look like:

```{r end_dplyr}
gapminder[gapminder$country == "Cambodia", c("year", "lifeExp")]
```

## Create a copy of `gapminder`

We're going to make changes to the `gapminder` tibble. To eliminate any fear that you're damaging the data that comes with the package, we create an explicit copy of `gapminder` for our experiments.

```{r}
(my_gap <- gapminder)
```

**Pay close attention** to when we evaluate statements but let the output just print to screen:

```{r eval = FALSE}
## let output print to screen, but do not store
my_gap %>% filter(country == "Canada")
```

... versus when we assign the output to an object, possibly overwriting an existing object.

```{r eval = FALSE}
## store the output as an R object
my_precious <- my_gap %>% filter(country == "Canada")
```

## Use `mutate()` to add new variables

Imagine we wanted to recover each country's GDP. After all, the Gapminder data has a variable for population and GDP per capita. Let's multiply them together.

`mutate()` is a function that defines and inserts new variables into a tibble. You can refer to existing variables by name.

```{r, live=TRUE}
my_gap %>%
  mutate(gdp = pop * gdpPercap)
```

Hmmmm ... those GDP numbers are almost uselessly large and abstract. Consider the advice of Randall Munroe of xkcd:

> One thing that bothers me is large numbers presented without context... 'If I added a zero to this number, would the sentence containing it mean something different to me?' If the answer is 'no,' maybe the number has no business being in the sentence in the first place."

Probably expressing the GDP in billions is a better idea:

```{r, live=TRUE}
my_gap %>%
  mutate(gdp_billion = pop * gdpPercap / 1e9)
```

Suppose we want to add a few more new columns:

1.  one expressing the population in millions,
2.  and another one that multiplies life expectancy (lifeExp) by population (pop). I guess this value would be the total years expected to live by a population.

With `mutate()`, we can add several new variables at once:

```{r, eval=FALSE, live=TRUE}
# my_gap %>%
#  mutate(
    gdp_billion = pop * gdpPercap / 1e9,
    popMil = round(pop / 1e6, 1),
    total_years = pop * lifeExp
#  )
```

## Use `arrange()` to row-order data in a principled way

`arrange()` reorders the rows in a data frame. Imagine you wanted this data ordered by year then country, as opposed to by country then year.

```{r}
my_gap %>%
  arrange(year, country) %>% 
  head(3)
```

Or maybe you want just the data from 2007, sorted on life expectancy?

```{r}
my_gap %>%
  filter(year == 2007) %>%
  arrange(lifeExp)
```

Oh, you'd like to sort on life expectancy in **desc**ending order? Then use `desc()`.

```{r}
my_gap %>%
  filter(year == 2007) %>%
  arrange(desc(lifeExp))
```

I advise that your analyses NEVER rely on rows or variables being in a specific order. But it's still true that human beings write the code and the interactive development process can be much nicer if you reorder the rows of your data as you go along. Also, once you are preparing tables for human eyeballs, it is imperative that you step up and take control of row order.

## Use `rename()` to rename variables

Say you are not much of a [`camelCase`](https://en.wikipedia.org/wiki/Camel_case) person, but rather you are all about [`snake_case`](https://en.wikipedia.org/wiki/Snake_case). So you are vexed by the variable names used in this data set. Let's rename some variables!

```{r, eval=FALSE, live=TRUE}
my_gap %>%
  rename(
    life_exp = lifeExp,
    gdp_percap = gdpPercap
    )
```

The thing to remember here is: `new_variable = oldVariable`.

I did NOT assign the post-rename object back to `my_gap` because that would make the chunks in this tutorial harder to copy/paste and run out of order. In real life, I would probably assign this back to `my_gap`, in a data preparation script, and proceed with the new variable names.

## `select()` can rename and reposition variables

You've seen simple use of `select()`. There are two tricks you might enjoy:

1.  `select()` can rename "on the fly" the variables you request to keep.
2.  `select()` can be used with `everything()` to hoist a variable up to the front of the tibble.

Filter data for Burundi after 1996, select `year`, `lifeExp`, and `gdpPercap` columns, rename `year` to `yr`, and move `gdpPercap` at the beginning:

```{r, eval=FALSE, live=TRUE}
my_gap %>%
  filter(country == "Burundi", year > 1996) %>%
  select(yr = year, lifeExp, gdpPercap) %>%
  select(gdpPercap, everything())
```

`everything()` is one of several helpers for variable selection. Read its help to see the rest.

## `group_by()` is a mighty weapon

I have found ~~friends and family~~ collaborators love to ask seemingly innocuous questions like, "which country experienced the sharpest 5-year drop in life expectancy?". In fact, that is a totally natural question to ask. But if you are using a language that doesn't know about data, it's an incredibly annoying question to answer.

dplyr offers powerful tools to solve this class of problem:

-   `group_by()` adds extra structure to your dataset -- grouping information -- which lays the groundwork for computations within the groups.
-   `summarize()` takes a dataset with $n$ observations, computes requested summaries, and returns a dataset with 1 observation.
-   Window functions take a dataset with $n$ observations and return a dataset with $n$ observations.
-   `mutate()` and `summarize()` will honor groups.

Combined with the verbs you already know, these new tools allow you to solve an extremely diverse set of problems with relative ease.

### Counting things up

Let's start with simple counting. How many observations do we have per continent?

```{r, eval=FALSE, live=TRUE}
my_gap %>%
  group_by(continent) %>%
  summarize(n = n())
```

Let us pause here to think about the tidyverse. You could get these same frequencies using `table()` from base R.

```{r}
table(gapminder$continent)
str(table(gapminder$continent))
```

But the object of class `table` that is returned makes downstream computation a bit fiddlier than you'd like. For example, it's too bad the continent levels come back only as *names* and not as a proper factor, with the original set of levels. This is an example of how the tidyverse smooths transitions where you want the output of step `i` to become the input of step `i + 1`.

The `count()` function is a convenient function that does both grouping and counting.

```{r}
my_gap %>%
  count(continent)
```

What if we wanted to add the number of unique countries for each continent? You can compute multiple summaries inside `summarize()`. Use the `n_distinct()` function to count the number of distinct countries within each continent.

```{r, eval=FALSE, live=TRUE}
my_gap %>%
  group_by(continent) %>%
  summarize(
  n = n(),
  n_countries = n_distinct(country)
  )
```

### General summarization

The functions you'll apply within `summarize()` include classical statistical summaries, like `mean()`, `median()`, `var()`, `sd()`, `mad()`, `IQR()`, `min()`, and `max()`. Remember they are functions that take $n$ inputs and distill them down into 1 output.

Although this may be statistically ill-advised, let's compute the average life expectancy by continent.

```{r}
my_gap %>%
  group_by(continent) %>%
  summarize(avg_lifeExp = mean(lifeExp))
```

What if we wanted to summarize multiple columns? We combine `summarize()` with `across()` to apply the same summary function(s) to multiple variables. Let's compute average and median life expectancy and GDP per capita by continent by year, but only for 1952 and 2007.

```{r, live=TRUE}
my_gap %>%
  filter(year %in% c(1952, 2007)) %>%
  group_by(continent, year) %>%
  summarize(across(c(lifeExp, gdpPercap), list(mean = ~ mean(.), median = ~ median(.))))
```

Let's focus just on Asia. What are the minimum and maximum life expectancies seen by year?

```{r, live=TRUE}
my_gap %>%
  filter(continent == "Asia") %>%
  group_by(year) %>%
  summarize(min_lifeExp = min(lifeExp), max_lifeExp = max(lifeExp))
```

## Grouped mutate

Sometimes you don't want to collapse the $n$ rows for each group into one row. You want to keep your groups, but compute within them.

### Computing with group-wise summaries

Let's make a new variable that is the years of life expectancy gained (lost) relative to 1952, for each individual country. We group by country and use `mutate()` to make a new variable. The `first()` function extracts the first value from a vector. Notice that `first()` is operating on the vector of life expectancies *within each country group*.

```{r, live=TRUE}
my_gap %>%
  group_by(country) %>%
  select(country, year, lifeExp) %>%
  mutate(lifeExp_gain = lifeExp - first(lifeExp)) %>%
  filter(year < 1963)
```

Within country, we take the difference between life expectancy in year $i$ and life expectancy in 1952. Therefore we always see zeroes for 1952 and, for most countries, a sequence of positive and increasing numbers.

Another useful function that you can use inside a `mutate` call is `lag()`, which finds the previous value in a vector.

# Assignment

So let's answer that "simple" question: which country experienced the sharpest 5-year drop in life expectancy?

Hints:

-   recall that this excerpt of the Gapminder data only has data every five years, e.g. for 1952, 1957, etc. So this really means looking at life expectancy changes between adjacent timepoints
-   the `top_n()` function may be useful.

```{r, live=TRUE}
my_gap %>% 
  group_by(country) %>%
  select(country, year, lifeExp) %>%
  mutate(LEchange = lifeExp - lag(lifeExp)) %>%
  arrange(LEchange)
```

------------------------------------------------------------------------

# Resources

-   dplyr official stuff: Package home [on CRAN](https://cran.r-project.org/package=dplyr); [introduction vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html)

-   Simple aggregation with the tidyverse: `dplyr::count()` and `dplyr::group_by()` + `dplyr::summarize()`, [Data transformation](http://r4ds.had.co.nz/transform.html) chapter in R for Data Science.

-   General aggregation with the tidyverse: [STAT 545 coverage](http://stat545.com/block024_group-nest-split-map.html) of general Split-Apply-Combine via nested data frames.

-   Simple aggregation with base R: `aggregate()`.

-   General aggregation with base R: `tapply()`, `split()`, `by()`, etc.

# License

These handouts are available under the [Creative Commons - Attribution-ShareAlike 4.0 International Licence](https://creativecommons.org/licenses/by-sa/4.0/) and were adapted from [Jenny Bryan's STAT 545](https://stat545.com/) available under the same license.
