---
title: 'R Graphics - practical'
author: "Marco Chierici, Giuseppe Jurman"
date: "Mar 14, 2023"
output:
  html_document:
    theme: readable
    toc: yes
    toc_float: yes
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60),
                      tidy = TRUE,
                      warning = FALSE,
                      error = TRUE,
                      collapse = TRUE)
set.seed(4561)
```

# Base R graphics

## Scatter plots

We begin by creating a data set consisting of 100 undergraduate students' math and reading test scores. The test scores are on a scale of 0 to 100. Each individual has also been assigned to either the Paper Test or the Electronic Test format (`TestFormat` column) and either the Classroom or Home setting (`TestLocation` column).

```{r data-generation}
set.seed(100)
MathGrade <- rnorm(n = 100, mean = 70, sd = 10)
set.seed(1000)
ReadingGrade <- rnorm(n = 100, mean = 65, sd = 13)
TestLocation <- c(rep("Classroom", 50), rep("Home", 50))
TestFormat <- c(rep("Paper", 25), rep("Electronic", 25), rep("Paper", 25), rep("Electronic", 25))
students <- tibble::tibble(MathGrade, ReadingGrade, TestLocation, TestFormat)

# Marginal Conditions
PaperTest <- students %>% dplyr::filter(TestFormat == "Paper")
ElectronicTest <- students %>% dplyr::filter(TestFormat == "Electronic")
Classroom <- students %>% dplyr::filter(TestLocation == "Classroom")
Home <- students %>% dplyr::filter(TestLocation == "Home")

# Cell Conditions
PaperTestHome <- PaperTest %>% dplyr::filter(TestLocation == "Home")
PaperTestClassroom <- PaperTest %>% dplyr::filter(TestLocation == "Classroom")
ElectronicTestHome <- ElectronicTest %>% dplyr::filter(TestLocation == "Home")
ElectronicTestClassroom <- ElectronicTest %>% dplyr::filter(TestLocation == "Classroom")
```


```{r, live=TRUE}
plot(students$MathGrade, students$ReadingGrade)
# equivalent with formula syntax:
plot(ReadingGrade ~ MathGrade, data=students)
```


```{r, live=TRUE}
# add a title
plot(students$MathGrade, students$ReadingGrade, main = "Title")
# and a subtitle
plot(students$MathGrade,
     students$ReadingGrade,
     main = "Title",
     sub = "Subtitle")
# add axis labels
plot(students$MathGrade, students$ReadingGrade, main = "Title", sub = "Subtitle", xlab = "Math grade", ylab = "Reading grade")
# change axis limits
plot(students$MathGrade, students$ReadingGrade, main = "Title", sub = "Subtitle", xlab = "Math grade", ylab = "Reading grade", xlim=c(0, 100), ylim=c(0, 100))
# put all together, and also remove the outer frame
plot(students$MathGrade, students$ReadingGrade, main = "Math grade vs. Reading grade", sub = "All conditions", xlab = "Math grade", ylab = "Reading grade", xlim=c(0, 100), ylim=c(0, 100), frame.plot = FALSE)
```

The function `scatterplot()` in the `car` package allows you to make enhanced scatter plots, with box plots in the margins, a regression line, a non-parametric regression smooth, and more.

```{r}
car::scatterplot(ReadingGrade ~ MathGrade, data=students)
```

This plot contains:

-   the data points
-   the regression line (solid)
-   the non-parametric regression smooth (dashed)
-   variability intervals (filled area)

The default non-parametric regression smooth is computed by local polynomial regression fitting (loess) with degree=1. We can customize the smoother's parameters using the `smooth=` argument: for example, let's fit a loess with degree=2.

```{r}
car::scatterplot(ReadingGrade ~ MathGrade, data=students,
                 smooth=list(degree=2))
```


Let's go back to the default scatterplot. We can now plot the data based on the different groups we created earlier. The following four plots show the Math vs. Reading scatterplots for each of the marginal groups: Paper Test, Electronic Test, Classroom location, and Home location.

```{r}
main_title <- "Math grade vs. Reading grade"
xlab <- "Math grade"
ylab <- "Reading grade"

plot(PaperTest$MathGrade, PaperTest$ReadingGrade, main=main_title,
     sub="Paper Test",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

plot(ElectronicTest$MathGrade, ElectronicTest$ReadingGrade, main=main_title,
     sub="Electronic Test",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

plot(Classroom$MathGrade, Classroom$ReadingGrade, main=main_title,
     sub="Classroom",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

plot(Home$MathGrade, Home$ReadingGrade, main=main_title,
     sub="Home",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))
```

We could plot the opposing situations (Paper and Electronic; Classroom and Home) side by side by setting the number of plots on the screen with the `par()` function, which affects the global graphical parameters. If we want to restore the parameters to their defaults, we have to save the "old parameters" while we set the new ones:

```{r, live=TRUE}
### Panel with PaperTest (left) and ElectronicTest (right)
# save the old parameters & set new ones
op <- par(mfrow=c(1, 1)) # create a plot with graphs arranged on 1 row and 2 columns

# plot(PaperTest$MathGrade, PaperTest$ReadingGrade, main=main_title,
#      sub="Paper Test",
#      xlab=xlab, ylab=ylab,
#      xlim=c(0, 100), ylim=c(0, 100))
plot(PaperTest$MathGrade, PaperTest$ReadingGrade, main=main_title,
     sub="Paper Test",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

# plot(ElectronicTest$MathGrade, ElectronicTest$ReadingGrade, main=main_title,
#      sub="Electronic Test",
#      xlab=xlab, ylab=ylab,
#      xlim=c(0, 100), ylim=c(0, 100))
plot(ElectronicTest$MathGrade, ElectronicTest$ReadingGrade, main=main_title,
     sub="Electronic Test",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

# reset the parameters
par(op)
```

```{r}
### Panel with Classroom (left) and Home (right)
# save the old parameters & set new ones
op <- par(mfrow=c(1, 2)) # create a plot with graphs on 1 row and 2 columns
plot(Classroom$MathGrade, Classroom$ReadingGrade, main=main_title,
     sub="Classroom",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

plot(Home$MathGrade, Home$ReadingGrade, main=main_title,
     sub="Home",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))
# reset the parameters
par(op)
```

We can also overplot different conditions on the same graph using the `points` function: to differentiate between conditions, we can use `col=` to change the points colors and `pch=` to change their shapes.

```{r}
# 1st plot
plot(PaperTest$MathGrade, PaperTest$ReadingGrade, main=main_title,
     sub="Paper Test",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

# points() just adds points to an existing plot:
# it inherits main, sub, labels, axes limits, etc. from the "parent" plot
points(ElectronicTest$MathGrade, ElectronicTest$ReadingGrade, main=main_title,
       pch=2, col="blue")
```

Time to add a legend to our plot! We use the `legend()` functions once we are done plotting. Note that:

-   We have to specify colors and shapes manually;
-   The legend's position can be one of topleft, topright, bottomleft, and bottomright.

```{r, live=TRUE}
plot(PaperTest$MathGrade, PaperTest$ReadingGrade, main=main_title,
     sub="Paper Test",
     xlab=xlab, ylab=ylab,
     xlim=c(0, 100), ylim=c(0, 100))

points(ElectronicTest$MathGrade, ElectronicTest$ReadingGrade, main=main_title,
       pch=2, col="blue")
legend("topleft", legend=c("Paper Test", "Electronic Test"), col=c("Black", "Blue"),
       pch=c(1, 2))
```

## Scatter plot matrices

Basic scatter plot matrix with `pairs()` on the `iris` dataset:

```{r, live=TRUE}
X <- iris %>% dplyr::select(-Species)
pairs(X, pch=19)
```

Show only upper panel:

```{r, live=TRUE}
pairs(X, pch=19, lower.panel=NULL)
```

Color points by groups (`Species` column):

```{r}
# define custom colors in RGB format
my_cols <- c("#00AFBB", "#E7B800", "#FC4E07")
```

```{r, live=TRUE}
# pairs(X, pch=19, lower.panel=NULL,
#       cex=0.5, # make the points smaller
#       col=---)
pairs(X, pch=19, lower.panel=NULL,
      cex=0.5, # make the points smaller
      col=my_cols[iris$Species])
```

## Box plots

In the following, we will look at data from the classical 1879 experiment of Michelson to measure the speed of light. This dataset is available in the `morley` object.

```{r}
(mm <- as_tibble(morley))
```

There are five experiments (column `Expt`) with 20 runs each (column `Run`); `sl` is the recorded speed of light, suitably coded.

Change `Expt` and `Run` into factors.

```{r}
mm$Expt <- factor(mm$Expt)
mm$Run <- factor(mm$Run)
```

Compare the five experiments with simple boxplots:

```{r, fig.height=8, fig.width=8, live=TRUE}
plot(Speed ~ Expt, data=mm, main="Speed of Light Data", xlab="Experiment No.")
```

Note that since `Expt` is a factor and `Speed` is numerical, `plot()` automatically draws boxplots for each factor level. We can also use the function `boxplot()`, as in the following examples.

```{r, live=TRUE}
# single boxplot
boxplot(mm$Speed)
# same as the previous boxplot comparing the five experiments, without plot frames
boxplot(Speed ~ Expt, data=mm, frame=FALSE)
# don't plot outliers
boxplot(Speed ~ Expt, data=mm, frame=FALSE, outline=FALSE)
# add some labels
boxplot(Speed ~ Expt, data=mm, frame=FALSE, outline=FALSE, main="Michelson Speed of light data", xlab="Experiment")
# horizontal boxplots
boxplot(Speed ~ Expt, data=mm, frame=FALSE, horizontal=TRUE)
# customizing colors
boxplot(Speed ~ Expt, data=mm, frame=FALSE, col=NULL, border="darkgreen")
```

## Strip charts

A strip chart (`stripchart()`) can be thought of a "1D scatterplot". Let's see a few examples with Michelson's Speed of light data set:

```{r}
stripchart(Speed ~ Expt, data=mm, pch=19)
# vertical plot with jittering
stripchart(Speed ~ Expt, data=mm, pch=19,
           vertical=TRUE, method="jitter")
# change point shapes and color by group
stripchart(Speed ~ Expt, data=mm, pch=1:5,  col=1:5,
           vertical=TRUE, method="jitter",
           main="Speed by Experiment",
           xlab="Experiment")

```

## Bar plots

We use here a subset of the `VADeaths` data set, representing the death rates in Virginia stratified by population subgroups:

```{r}
VADeaths

# consider only the 1st three rows, to simplify
x <- VADeaths[1:3, "Rural Male"]
```

```{r, live=TRUE}
# basic bar plot
barplot(x)
# horizontal version
barplot(x, horiz=TRUE)
# change group names
barplot(x, names.arg=c("A", "B", "C"))
# customize colors
barplot(x, col="white", border="steelblue")
```

```{r}
# different colors for each group
barplot(x, col = c("#999999", "#E69F00", "#56B4E9"))
# add labels
barplot(x, col = c("#999999", "#E69F00", "#56B4E9"),
        main="Death rates in Virginia",
        xlab="Age group",
        ylab="Rate")
```

Bar plots can also be stacked or grouped:

```{r}
# define some colors
palette <- ggsci::pal_startrek()
my_cols <- palette(5)

# stacked bar plots
barplot(VADeaths, legend=rownames(VADeaths), col=my_cols)

# grouped bar plots
barplot(VADeaths, legend=rownames(VADeaths), col=my_cols,
        beside=TRUE)

# customized legend
barplot(VADeaths, beside=TRUE, col=my_cols)
legend("topleft", legend=rownames(VADeaths), fill=my_cols,
       box.lty=0, cex=0.8)

```

## Line plots

Basic line plots are drawn using a combination of `plot(x, y, type="l")` and `lines(x, y)`. We generate some variables and see a couple of examples:

```{r}
# data generation
x <- seq(1, 10)
y1 <- x*x
y2 <- 2*y1

# stair steps plot
plot(x, y1, type="S",
     xlab="x", ylab="y")

# points + lines, custom color
plot(x, y1, type="b", pch=19, col="darkorange",
     xlab="x", ylab="y")

# multiple lines:
# 1. draw the 1st line plot
plot(x, y1, type="b", pch=19, col="blue",
     xlab="x", ylab="y")
# 2. add the 2nd line
lines(x, y2, pch=18, type="b", col="darkred", lty=2)
# 3. (optional) add a legend
legend("topleft", legend=c("x^2", "2x^2"), col=c("blue", "darkred"), lty=1:2, lwd=2, cex=0.8)
```

## Histograms and density plots

Let's go back to our student's scores data set that we created initially and draw a histogram of the `MathGrade` values.

A basic histogram of a numerical vector `x` is created by the function `hist(x, [breaks=b])`, optionally specifying the (number of) breakpoints for binning `x` values.

```{r, live=TRUE}
x <- students$MathGrade

# basic histograms
hist(x)
hist(x, breaks=5)
hist(x, breaks=20)
# custom color
hist(x, col="steelblue")
```

To produce a density plot, we first use `density()` to estimate the kernel density, which we can then draw with a standard call to `plot()`:

```{r, live=TRUE}
dens <- density(x)
plot(dens, col="blue", main="Density of Math grades")
# a filled version using polygon():
polygon(dens, col="blue")
```

## QQ plots (rankit plots)

Quantile-quantile plots are useful to visually compare a data set with the normal distribution. A typical scenario involves fitting a linear regression model to some data and then checking the residuals for skewness, kurtosis and outliers (i.e., is the error term actually normally distributed?).

```{r}
# load the 'faithful' data set
data(faithful)
x <- as_tibble(faithful)
x
```

We fit a linear model to predict the duration of an eruption (`eruptions`) from the waiting time between eruptions (`waiting`) for the Old Faithful geyser in Yellowstone National Park.

```{r}
lm_fit <- lm(eruptions ~ waiting, data=x)
summary(lm_fit)
```

Then we check whether the residuals follow a normal distribution.

```{r, live=TRUE}
qqnorm(resid(lm_fit), main="Residuals rankit plot")
qqline(resid(lm_fit))
```

## Dot charts

Dot charts (or Cleveland dot plots) are typically used to visually represent ranked data. We'll see an example on the `mtcars` dataset:

```{r}
as_tibble(mtcars)
```

Default dot chart:

```{r}
# arrange the data by increasing mpg
x <- mtcars %>% dplyr::arrange(mpg)
dotchart(x$mpg, labels=rownames(x), cex=0.5, pch=19, xlab="mpg")
```

Dot chart showing groups of observations with different colors:

```{r}
# group by "cyl" and color groups
grps <- as.factor(x$cyl)
# select the required number of colors from a custom palette
my_cols <- ggsci::pal_futurama()(nlevels(grps))
```

```{r, live=TRUE}
# create the dot chart
dotchart(x$mpg, labels=rownames(x),
         groups=grps, gcolor=my_cols,
         color=my_cols[grps],
         cex=0.6, pch=19, xlab="mpg")
```
